PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-21 papers/NAR_web_server_2009/paper.sup.tex
Should be the final version of all components of the resubmission.

\mc{Runtime of the tool is analyzed in Figure~4 of the main paper.  In
Figure~\ref{memory} we present a plot of the memory footprint of each tool
as a function of input size.  Data points were generated by the application of
each tool to simulated sets of a specified size (generated by ESTSim),
measuring the maximum amount of memory required for the run.  We see that while
all tools require memory linear in the input size, \wcd\/ requires
slightly less than does \peace, while both require significantly less
memory than \capthree.  In general we find the footprint of \peace\/
to be relativily small, as compared to other clustering tools
(discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).}

>>>>

\mc{Runtime of the tool is analyzed in Figure~4 of the main paper.  In
Figure~\ref{memory} we present a plot of the memory footprint of each tool
as a function of input size.  Data points were generated by the application of
each tool to simulated sets of a specified size (generated by ESTSim),
measuring the maximum amount of memory required for the run.  We see that while
all tools require memory linear in the input size, \wcd\/ requires
slightly less than does \peace, while both require significantly less
memory than \capthree.  In general we find the footprint of \peace\/
to be relativily small, as compared to other clustering tools
(discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).

All runs were conducted on 3.0 GHz Intel Xeon EM64T CPUs with 2 MB
caches and a 800 MHz front side bus, model number Xeon LV 3.0 (2005)}

### 2010-04-21 papers/NAR_web_server_2009/paper.tex
Should be the final version of all components of the resubmission.

\history{Received Feb. 13, 2010 \\ Resubmitted April 20, 2010}

@@

However, clustering is a computationally challenging problem.  Even
with the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster on the basis of pair-wise
sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS technologies exacerbates
this problem.  To deal with this, \peace\/ combines our own version of
the $d^2$ alignment-free sequence distance function \cite{Hide94} and
the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
and accurately find clusters of ESTs expressed from the same gene
without reference to a sequenced genome.  Compared against \wcd, the
leading clustering tool in the literature \cite{Hazelhurst08a}, as well as
other tools designed for the same purpose 
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
sensitive and more robust to sequencing error without sacrificing
runtime.  Nor are any of these tools designed for the ease of installation
and use that \peace\/ provides.

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For
our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
to generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~D.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  \mc{The most important method of quality
  assessment is {\it sensitivity} (the fraction of fragment pairs from
  the same gene that were} \mc{correctly clustered together).  But we
  also look at the {\it Jaccard Index} (which balances sensitivity
  with the number of false positives by looking at the ratio of true
  positives to the sum of true positives, false positives, and true
  negatives), {\it Type 1 error} (the fraction of genes that were
  divided between clusters), and {\it Type 2 error} (the fraction of
  clusters containing two or more genes).  Specificity was not an
  issue in simulated tests, as both \peace\/ and \wcd\/ were
  completely successful in separating unrelated ESTs (with the
  exception of recently duplicated genes, as discussed in the
  Supplementary Materials).  In Figure~\ref{SeJiT1T2} we plot these
  four tests as a function of error rate, and observe the almost
  identical results between \peace\/ and \wcd\/.  In
  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
  \wcd\/, again observing almost identical results when run
  sequentially -- but significantly faster runtime for PEACE on
  multiple processor when holding the EST/processor ratio constant
  (ranging from a 65\% improvement for two processors too a 17\%
  improvement for 12 processors).}

\mc{We further note, as a point of interest that, we can significantly
  improve the quality of \peace simulation results through the
  increase of the threshold value (see Supplementary Materials,
  Section C) -- achieving a significant improved in PEACE sensitivity
  without an adverse effect on the Jaccard index (i.e. a significant
  increase in the incorrect merging of clusters).  However, the
  improvement does not carry over to the application of real data,
  where we observe a significant drop in the Jaccard Index.  Due to
  this drop we have opted to leave the lower threshold value as the
  default, but users might consider the higher value, thus decreasing
  the risk of incorrectly splitting clusters, if planning on
  using on the subsequent use of a assembly tool that capable of
  breaking up the incorrectly merged clusters during transcript reconstruction.}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
  We notice essentially identical results for PEACE and WCD in
  quality, both significantly better than Cap3 in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with PEACE showing a
  60\% runtime improvement in the first data set, but requiring 20\%
  more time in the Arabidopsis data set.  In Section~D.1 of the
  Supplementary Materials we present runtimes several more sets,
  observing that while \peace\/ appears to be significantly 
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}

@@

\mc{We tested the three tools on short-read data using the \metasim\/
  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
  \metasim\/ are sequence generation and error models corresponding to
  several technologies, including the 454 short-read technology.
  Leaving \capthree\/ and \wcd\/ at their default values, but
  employing the {\it adaptive $d^2$} strategy for short read sequences
  (see Section~C in the supplementary materials), we find a that
  \peace\/ performs significantly better than the other two tools on
  454 data using the default \metasim\/ model (with sequence sizes
  ranging in size from 200bp to 370bp, at an average size of 250bp and
  a standard deviation of 17bp).  Over thirty runs, we see an average
  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
  \capthree.  For the Jaccard Index we get essentially the same
  numbers.  For 100 transcript \peace\/ creates 128 clusters (as
  opposed to 6,370 and 19245 clusters for \wcd\/ and \capthree), and
  creates only 3 singletons (as opposed to 5,850 and 15432).  We do
  note that it is likely that the \wcd results could be considerably
  improved by the adjustment of a few parameters, but were unable to
  determine how to do that.  (Runtime of \peace\/ was somewhat slower
  than that of \wcd, but given the significant improvement in quality
  of results this is unimportant.)}

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline EasyCluster Human & \peace & 0.871 & 0.809 &
\cline{2-9} Benchmark & \wcd &  0.020 & 0.020 &
\cline{2-9} (111 Genes) & \capthree & 0.005 & 0.004 &
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

\mc{Here we have presented \peace, a stand alone tool for the
  high-throughput clustering of transcript fragments capable of
  dealing with sequences as short as 50 bases.  \peace\/, available at
  \href{http://www.peace-tools.org}{www.peace-tools.org}, is
  open-source and managed through a user friendly GUI that enables
  both local and remote installation and execution in sequential or
  parallel mode.  Based on a novel minimum-spanning-tree based
  algorithm for the clustering of the fragments by gene association,
  \peace\/ shows significant improvement in sensitivity, over the
  competing \wcd\/ tool \cite{Hazelhurst08a} when applied to NGS
  reads, matches \wcd\/ when applied to Traditional Sanger Sequencing
  output, and shows an order of magnitude in improvement over the
  clustering performed in the course of assembly by the \capthree\/
  tool \cite{Huang99}.}

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, other natural biological effects (e.g. the trans-splicing of
transcripts), effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  The problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small cluster (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.

Peace can be downloaded from
\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
committed to maintaining and improving the tool.  The underlying PEACE
architecture is configured to support a variety of tasks, including the
possibility of a minimum spanning tree based assembly tool we hope to
add to the tool in a future study.

>>>>

\history{Received Feb. 13, 2010 \\ Resubmitted April 21, 2010}

@@

However, clustering is a computationally challenging problem.  Even
with the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster on the basis of pair-wise
sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS technologies exacerbates
this problem.  To deal with this, \peace\/ combines our own version of
the $d^2$ alignment-free sequence distance function \cite{Hide94} and
the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
and accurately find clusters of ESTs expressed from the same gene
without reference to a sequenced genome.  \mc{Compared against
  clustering tool in the literature
  \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
    Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
  sensitive and robust than the other tools, and more user friendly
  than any published clustering tool.  Nor are any of} \mc{these tools
  designed for the ease of installation and use that \peace\/
  provides.}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).
\mc{For our simulation tests we used the {\bf ESTSim} tool to generate
  simulated Sanger sequenced transcript fragments, and the \metasim\/
  tool to generate simulated short read sequences from 454 and Illumina
  technologies \cite{Hazelhurst03,Richter2008}.  (See Supplementary
  Materials, Section~D.1, for details on the sequence size and error
  models.)  Fragments were generated from the list of 100 zebra fish
  genes used in the \wcd\/ testing \cite{Hazelhurst08a}.  Tool
  parameters were taken to match, as closely as possible, those used
  in the \wcd\/ study (see Supplementary Materials).  }

\mc{The most important method of quality assessment is {\it
    sensitivity} (the fraction of fragment pairs from the same gene
  that were} \mc{correctly clustered together).  We also look at the
  {\it Jaccard Index} (which balances sensitivity with the number of
  false positives), {\it Type 1 error} (the fraction of genes that
  were divided between clusters), and {\it Type 2 error} (the fraction
  of clusters containing two or more genes)
  \cite{Hazelhurst08a,Wang04}.  In Figure~\ref{SeJiT1T2} we plot these
  four tests as a function of error rate, and observe the almost
  identical results between \peace\/ and \wcd\/.  In
  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
  \wcd\/, again observing almost identical results when run
  sequentially -- but significantly faster runtime for PEACE on
  multiple processor when holding the EST/processor ratio constant
  (ranging from a 65\% improvement for two processors to a 17\%
  improvement for 12 processors).}

\mc{In the course of our simulations we also investigated the memory
  footprint -- of particular concern in the large data sets commonly
  encountered.  In short, we find that, as with \wcd, the required
  memory for an execution is linear in the size of the data set --
  requiring slightly more memory than \wcd\/ and considerably less
  than \capthree.  We present relevant results in Section D.4 and
  Figure~S3 of the Supplementary Materials.}

% Appologizes to anyone working with this: the coloring would not
% presist between columns, and the break point kept changing on me.
% Modifying the \mc macro will allow for color removal without having
% to take all of this out.
\mc{We note, as a point of interest, that we can significantly improve
  the quality of \peace\/} \mc{simulation} \mc{results} \mc{through} \mc{the} \mc{increase} \mc{of}
  \mc{the} \mc{threshold} \mc{value} \mc{(see} \mc{Supplementary} \mc{Materials,} \mc{Section~D.5) --}
  \mc{achieving} \mc{a} \mc{significant} \mc{improvement} \mc{in PEACE sensitivity without an
  adverse effect on the Jaccard Index.  However, the improvement does
  not carry over to the application of real data (Supplementary
  Materials, Section E.2), where we observe a significant increase in
  the incorrect merging of clusters.  While this might be acceptable
  to a user planning to employ an assembly tool capable of breaking
  up the clusters (but unable to re-merge a split cluster), for the
  purposes of clustering assessment it makes sense base our analysis (and
  set out default values) at the lower threshold.}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section D, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and a 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
  \cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).
  We notice essentially identical results for \peace\/ and wcd\/ in
  quality, both significantly better than \capthree\/ in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with \peace\/ showing a
  60\% runtime improvement over \wcd\/ in the first data set, but requiring 20\%
  more time than \wcd\/ in the Arabidopsis data set.  (We cannot make
  a reasonable runtime comparison to \capthree, as the runtime of that
  tool reflects both clustering and assembly.)  In Table~S1 of the
  Supplementary Materials we present runtimes of several more sets,
  observing that while \peace\/ appears to be significantly 
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}

@@

\mc{We tested the three tools on short-read data using the \metasim\/
  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
  \metasim\/ are sequence generation and error models corresponding to
  several technologies, including the 454 short-read technology
  (producing reads ranging in size from 200bp to 370bp with a mean of
  250bp and a standard deviation of 17bp), and the Illumina short-read
  technology (producing reads of exactly 62bp).  Experiments were run
  using two sets of \wcd\/ parameters (the default parameters, and
  setting window size to 75 and threshold to 75 -- thus matching the
  equivalent \peace\/ parameters), and using default values
  of \capthree.  Only with \peace\/ were we able to get reasonable
  results, seeing a sensitivity of 0.871 for the 454 reads and
  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
  0.372 respectively, and other comparably figures.  (See Table~S3 in
  the Supplementary Materials.)  In short, \peace performs quite well
  for 454 reads and provides some useful information for Illumina
  reads, while \wcd\/ and \capthree\/ are unable to produce working
  results.}

@@

\mc{Here we have presented \peace, a stand alone tool for the
  high-throughput clustering of transcript fragments capable of
  dealing with sequences as short as 50 bases.  \peace\/ is
  open-source and managed through a user friendly GUI that enables
  both local and remote installation and execution in sequential or
  parallel mode.  Based on a novel minimum spanning tree based
  algorithm for the clustering of the fragments by gene association,
  \peace\/ shows significant improvement in sensitivity over the
  competing \wcd\/} \mc{tool} \cite{Hazelhurst08a} \mc{when} \mc{applied} \mc{to} \mc{NGS}
  \mc{reads,} \mc{matches} \mc{\wcd\/} \mc{when} \mc{applied} \mc{to} \mc{Sanger} \mc{sequencing}
  \mc{output,} \mc{and} \mc{shows} \mc{an} \mc{order} \mc{of} \mc{magnitude} \mc{in improvement over the
  clustering performed in the course of assembly by the \capthree\/
  tool \cite{Huang99}.}

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, other natural biological effects (e.g. the trans-splicing of
transcripts), effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  These problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small cluster (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.

\mc{Peace can be downloaded from
\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
committed to keep maintaining and improving the tool in the future.
Meanwhile, we are developing our own MST-based assembly tool that can
seamlessly integrate with PEACE.  The underlying modular design of
PEACE offers users many possibilities to expand and incorporate the
MST algorithm for other bioinformatics application.}
